Designing Qusecs Optimally: Quasi-uniform or Self-similar Layered Materials as Recursively Decomposed Solutions of Geometric Constraint Systems

### 2014-12-18 doc/GMP2015/s02_preliminaries.tex
Meera's changes to preliminaries.

\section{Preliminaries}
\label{sec:prelim}

\subsection{Basics}

In this paper, a \dfn{geometric constraint system} is a multivariate
polynomial (usually bilinear or quadratic) system $G(x,\delta)=0$,
representing constraints with parameters $\delta$ between geometric
primitives  in $\mathbb{R}^2$ represented collectively as $x\in
\mathbb{R}^n$.
%
When the type of constraint (system) is fixed, the system is simply
represented as $(G,\delta)$, where $G$ is the underlying constraint
(hyper)graph $G = (V,E)$ with the vertices $V$ representing the
geometric primitives in $\mathbb{R}^2$ and (hyper)edges $E$
representing the constraints, each with an associated parameter
$\delta$.
%
For example, a \dfn{bar-joint system or linkage} $(G,\delta)$, is a
graph $G=(V,E)$ with fixed length bars as edges, i.e. $\delta: E
\rightarrow \mathbb{R}$; this represents the distance constraint
system $\| x_u -x_v \|_2 = \delta_{u,v}$ for  $(u,v) \in E$, where
$x_u \in \mathbb{R}^2$ represents the coordinates of $u\in V$.

\noindent
\note Geometric constraint systems can also have
inequalities, where the parameters in $\delta$ are small intervals of
values rather than exact values.

@@

Other generic constraint system or (hyper)graph properties are the
following:
% (\note we refer to these as properties of the constraint
% system or as properties of the underlying (hyper)graph
% interchangeably).
A constraint (hyper)graph $G$ is \dfn{independent} if its appropriate
rigidity matrix of indeterminates has independent rows (i.e, the
determinant of some square submatrix is not identically zero).
%It is {\em rigid} if the number of independent rows of the rigidity matrix is
%at least the number of columns less the number of rigid body motions,
%which is 3 for distance constraint systems.
It is \dfn{isostatic (minimally rigid, well-constrained)} if it is
both rigid and independent.
%if the number of generically independent rows or the rank of the appropriate rigidity matrix is maximal.
%For example, the maximal rank of a bar-joint rigidity matrix is $2|V| - 3$,
%where $3$ is the number of rotational and translational degrees-of-freedom of a rigid body in $\mathbb{R}^2$.
%The graph $G$ is {\em rigid} if there exists some spanning subgraph $S\subseteq G$ such that $S$ is well-constrained.
It is \dfn{flexible} if it is not rigid, \dfn{under-constrained} if it is independent and not rigid, or \dfn{over-constrained} if it is not independent.

Defining the independence of a subset of edges $E' \subseteq E$ to be
the independence of corresponding rows in the appropriate rigidity
matrix of indeterminates, we obtain the \dfn{rigidity matroid} of a
constraint (hyper)graph $G = (V,E)$.
%The 2-dimensional {\em rigidity matroid} of a constraint (hyper)graph $G = (V,E)$ is a linear matroid  on $E$,
%where a subset of edges $E' \subseteq E$ is {\em independent} in the matroid,
%if the set of corresponding rows in the appropriate rigidity matrix of indeterminates are linearly independent.
There are various results on combinatorial characterization of
independence and rigidity for different types of  rigidity matroid.
For bar-joint rigidity matroid, the famous Laman's theorem \uncited
states that the underlying graph is minimally rigid if and only if
$|E| = 2|V|-3$ and $|E'| \le 2|V'|-3$ for every induced subgraph with
at least 2 vertices. The result by Lovasz and Yemini \uncited shows
that all 6-vertex-connected graphs are rigid in the plane. For bar-
body rigidity matroid, Tay \uncited proved that the underlying
multigraph is minimally rigid if and only if it can be decomposed as
$3$ edge disjoint spanning trees. White and Whiteley \uncited gave
the same characterization using a different technique to study the
algebraic - geomtric conditions called pure condition. Lee, Streinu
and Theran \uncited defined the \dfn{$(k,l)$-sparsity matroid},
where a hypergraph $G$ is called \dfn{$(k,l)$-sparse} if $|E'| \le
k|V'| - l$ for any induced subgraph $(V',E')$ with at least 2
vertices, and \dfn{$(k,l)$-tight} if is $(k,l)$-sparse and $|E| = k|V|
- l$. In general, given a $d$-uniform hypergraph, a $(k,l)$-sparsity
condition is matroidal as long as $l \le dk-1$.

In this paper, a \dfn{qusec} is any \dfn{independent} geometric
constraint system of one of the 3 types mentioned above. In the
remainder of this section and the next two sections~\ref{sec:DRP} and
\ref{sec:recomb} we only consider bar-joint qusecs and graphs.
Relevant formal analogies for the other 2 types of qusecs and
(hyper)graphs are given in the subsequent 2 sections on Applications.

@@

@@

% \begin{definition}
%     A \textbf{trivial} graph is ill-defined in the general case. The only strict requirements are: (1) it must be an over-constrained graph and (2) all of its subgraphs are also trivial.
%     \todo{Maybe leave the following out until it's needed later?}
%     In the familiar geometric cases of $d$-dimensional space, all vertex weights are $d$, all edge weights are $1$, and constant $k= -{{d+1}\choose{2}}$. Trivial graphs for 2D would be a single vertex. Trivial graphs in 3D would be a vertex and an edge (2 vertices with an edge between). Etc. These trivial graphs capture the notion of the rotational symmetry that exists in geometric spaces.
% \end{definition}

A \dfn{trivial} graph is a single vertex.
%
\begin{definition}\label{def:drp}
    The \dfn{decomposition-recombination \mbox{(DR-)} plan} of graph
    $G$, \uncited $DRP(G)$, is defined as a forest that has the
    following properties:
    \begin{enumerate}
        \item Each node represents/contains/is a rigid vertex-induced
        subgraph of $G$.
        % \item For a node, $C$, that is  a non-trivial graph, its $N$
        % children, $C_1, \ldots, C_N$, are rigid vertex-maximal proper
        % subgraphs of $C$.
        \item The vertex set of $\bigcup_{i=1}^N{C_i}$ is the vertex
        set of $C$.
        \item A leaf node is a trivial subgraph.
        \item A root node is a maximal rigid subgraph of $G$.
    \end{enumerate}

    % It can also be described recursively as: the root is $G$, its children are the trivial subgraphs and the DR-plans of its well-constrained vertex-maximal proper subgraphs whose union is $G$ itself.

    An equivalent DAG can be constructed from the above forest such
    that all nodes containing the same subgraphs are identified into a
    single node. In this manuscript, however, we retain a forest as
    defined above. When a graph is isostatic, its DR-plan is a tree.

    %Note that nodes will be referred to interchangeably as ``the node that
    %represents or contains the (sub)graph $C$'' and as simply ``$C$''.

    A DR-plan is \dfn{complete} if it satisfies an additional rule:
        \item For a node, $C$, that is  a non-trivial graph, its $N$
        children, $C_1, \ldots, C_N$, are all of the rigid vertex-
        maximal proper subgraphs of $C$. This makes Rule 2 implicit,
        and in fact, $C$ contains all the edges in the union of its
        children as well. The complete DR-plan is unique.
    A DR-plan is \dfn{optimal} if it minimizes the maximum fan-in over all nodes in the tree. It is not necessarily unique.
\end{definition}

@@

\noindent
\note In Section \ref{sec:DRP} $G$ is assumed to be isostatic.
$Idc(G,X)$ is the subgraph of $G$ induced by the vertex set
$X\subseteq V$. For a subgraph $H=(W,F)$ of $G$, $Idc(G,H)=Idc(G,W)$.
In general, $C$ is the graph in an arbitrary node in $CompleteDRP(G)$.
$C_i$ is the $i^{\text{th}}$ child of $C$. It is implied that $C_i$ is
an isostatic vertex-maximal proper subgraph.

\subsection{Importance of an Optimal DR-plan}
The use of a DR-plan in designing especially self-similar qusecs is
immediately clear, as shown in Figures \ref{XX} and \ref{XX}
\todo{Reference?}. In fact, knowing an optimal DR-plan is crucial for
design and analysis of even non-self-similar qusecs and their
realizations with various desirable properties.

First, note that the basic task of \header{realizing given qusecs}
would require finding real solutions a large multivariate polynomial
system (of inequalities and equalities representing the constraints).
Without DR-planning, this requires double exponential time in the
number of variables (even if orientation type is specified). With DR-
planning, the complexity is dominated by the size of the largest
subsystem that is solved, or recombined from the solutions of its
child subsystems i.e.,the maximum fan-in occuring in a DR-plan. Thus
finding an optimal DR-plan is important for realization.

Next, using the DR-plan we obtain a \header{recursive block
decomposition of the rigidity matrix} with each block being the
rigidity matrix of an isostatic subgraph. The blocks intersect on
isostatic or trival subgraphs. This recursive block decomposition of
the rigidity matrix $R$ correspondingly \header{recursively decomposes
the stress vectors} $s$ that balance a given external stress $t$ since
$sR = t$, and moreover, the external stress vector $t$ now
\header{recursively ``distributes'' as external stresses} for each
isostatic subgraph of the DR-plan. When the rigidity matrix and stress
vectors are indeterminates, it can be used to design  qusecs
underlying graph and parameters $\delta$ to obtain an optimal
distribution of stresses on the geometric primitives. When the
realization of the qusec is given, this can be used to analyze the
distribution of a given external load.

>>>>

\section{Preliminaries}
\label{sec:prelim}
%
\subsection{Basics}
%
In this paper, a \dfn{geometric constraint system} is a multivariate
polynomial (usually bilinear or quadratic) system $G(x,\delta)=0$,
representing constraints with parameters $\delta$ between geometric
primitives  in $\mathbb{R}^2$ represented collectively as $x\in
\mathbb{R}^n$.
%
When the type of constraint (system) is fixed, the system is simply
represented as $(G,\delta)$, where $G$ is the underlying constraint
(hyper)graph $G = (V,E)$ with the vertices $V$ representing the
geometric primitives in $\mathbb{R}^2$ and (hyper)edges $E$
representing the constraints, each with an associated parameter
$\delta$.
%
For example, a \dfn{bar-joint system or linkage} $(G,\delta)$, is a
graph $G=(V,E)$ with fixed length bars as edges, i.e. $\delta: E
\rightarrow \mathbb{R}$; this represents the distance constraint
system $\| x_u -x_v \|_2 = \delta_{u,v}$ for  $(u,v) \in E$, where
$x_u \in \mathbb{R}^2$ represents the coordinates of $u\in V$.
%
\noindent
\note Geometric constraint systems can also have
inequalities in addition to equations, where the parameters in $\delta$ are small intervals of
values rather than exact values.

@@

Other generic constraint system or (hyper)graph properties are the
following:
% (\note we refer to these as properties of the constraint
% system or as properties of the underlying (hyper)graph
% interchangeably).
A constraint (hyper)graph $G$ is \dfn{independent} if its appropriate
rigidity matrix of indeterminates has independent rows (i.e, the
determinant of some square submatrix is not identically zero).
%It is {\em rigid} if the number of independent rows of the rigidity matrix is
%at least the number of columns less the number of rigid body motions,
%which is 3 for distance constraint systems.
It is \dfn{isostatic (minimally rigid, wellconstrained)} if it is
both rigid and independent.
%if the number of generically independent rows or the rank of the appropriate rigidity matrix is maximal.
%For example, the maximal rank of a bar-joint rigidity matrix is $2|V| - 3$,
%where $3$ is the number of rotational and translational degrees-of-freedom of a rigid body in $\mathbb{R}^2$.
%The graph $G$ is {\em rigid} if there exists some spanning subgraph $S\subseteq G$ such that $S$ is well-constrained.
It is \dfn{flexible} if it is not rigid, \dfn{underconstrained} if it
is independent and not rigid, or \dfn{overconstrained} if it is not
independent.

Defining the combinatorial independence of a subset of edges $E'
\subseteq E$ to be the generic independence of corresponding rows in
the appropriate rigidity matrix of indeterminates, we obtain the
\dfn{rigidity matroid} of a constraint (hyper)graph $G = (V,E)$.%The 2-dimensional {\em rigidity matroid} of a constraint (hyper)graph $G = (V,E)$ is a linear matroid  on $E$,
%where a subset of edges $E' \subseteq E$ is {\em independent} in the matroid,
%if the set of corresponding rows in the appropriate rigidity matrix of indeterminates are linearly independent.
There are various results on combinatorial characterization of
independence and rigidity  and rigidity matroids for different types
of graphs. For bar-joint rigidity matroid, the famous Laman's theorem
\uncited states that the underlying graph is isostatic if and only if
$|E| = 2|V|-3$ and $|E'| \le 2|V'|-3$ for every induced subgraph with
at least 2 vertices. The result by Lovasz and Yemini \uncited  shows
that all 6-vertex-connected graphs are rigid in the plane. For bar-
body rigidity matroid, Tay \uncited proved that the underlying
multigraph is isostatic if and only if it can be decomposed as $3$
edge disjoint spanning trees. White and Whiteley \uncited gave the
same characterization using a different technique to study the
algebraic-geometric conditions of genericity, called pure condition.
Lee, Streinu and Theran \uncited defined the \dfn{$(k,l)$-sparsity
matroid}, where a hypergraph $G$ is called \dfn{$(k,l)$-sparse} if
$|E'| \le k|V'| - l$ for any induced subgraph $(V',E')$ with at least
2 vertices, and \dfn{$(k,l)$-tight} if is $(k,l)$-sparse and $|E| =
k|V| - l$. In general, given a $d$-uniform hypergraph, a
$(k,l)$-sparsity condition is matroidal as long as $l \le dk-1$.

In this paper, a \dfn{qusec} is any \dfn{independent} geometric
constraint system of one of the 3 types mentioned above.
\medskip\noindent
\note In the remainder of this section and the next three
sections~\ref{sec:DRP} and \ref{sec:recomb} we only consider bar-joint
qusecs and graphs. Relevant formal analogies for the other 2 types of
qusecs and (hyper)graphs are given in the subsequent 2 sections on
Applications.

@@

The notation $Idc(G,X)$ is the subgraph of $G$ induced by the vertex set
$X\subseteq V$. For a subgraph $H=(W,F)$ of $G$, $Idc(G,H)=Idc(G,W)$.

@@

% \begin{definition}
%     A \textbf{trivial} graph is ill-defined in the general case. The only strict requirements are: (1) it must be an over-constrained graph and (2) all of its subgraphs are also trivial.
%     \todo{Maybe leave the following out until it's needed later?}
%     In the familiar geometric cases of $d$-dimensional space, all vertex weights are $d$, all edge weights are $1$, and constant $k= -{{d+1}\choose{2}}$. Trivial graphs for 2D would be a single vertex. Trivial graphs in 3D would be a vertex and an edge (2 vertices with an edge between). Etc. These trivial graphs capture the notion of the rotational symmetry that exists in geometric spaces.
% \end{definition}
%
\begin{definition}\label{def:drp}
    The \dfn{decomposition-recombination \mbox{(DR-)} plan} of graph
    $G$, \uncited {\sl DRP}$(G)$ (see Figure \ref{fig:c2c3ofk33s}, is defined as a forest that has the
    following properties:
    \begin{enumerate}
        \item Each node represents/contains/is a rigid vertex-induced
        subgraph of $G$.
        % \item For a node, $C$, that is  a non-trivial graph, its $N$
        % children, $C_1, \ldots, C_N$, are rigid vertex-maximal proper
        % subgraphs of $C$.
        \item The vertex set of $\bigcup_{i=1}^N{C_i}$ is the vertex
        set of $C$.
        \item A leaf node is a single vertex.
A \dfn{trivial} graph is empty or a single vertex.
        \item A root node is a maximal rigid subgraph of $G$.
    \end{enumerate}
%
    % It can also be described recursively as: the root is $G$, its children are the trivial subgraphs and the DR-plans of its well-constrained vertex-maximal proper subgraphs whose union is $G$ itself.

    A DR-plan is \dfn{complete} if it satisfies an additional rule:
        \item For a node $C$ that is  a non-trivial graph, its $N$
        children, $C_1, \ldots, C_N$, are all of the rigid vertex-
        maximal proper subgraphs of $C$. This makes Rule 2 implicit,
        and in fact, $C$ contains all the edges in the union of its
        children as well.
    A DR-plan is \dfn{optimal} if it minimizes the maximum fan-in over all nodes in the tree.
    The maximum  fan-in is called the \dfn{size} of the DR-plan

%In general, $C$ is the graph in an arbitrary node in $CompleteDRP(G)$.
%$C_i$ is the $i^{\text{th}}$ child of $C$. It is implied that $C_i$ is
%an isostatic vertex-maximal proper subgraph.
    %Note that nodes will be referred to interchangeably as ``the node that
    %represents or contains the (sub)graph $C$'' and as simply ``$C$''.
\end{definition}
%
% \todo{This was labeled as both begin(remark) and end(observation), what was your intent?}
\begin{remark}
    For a given graph, there could be exponentially many  DR-plans and
    even optimal DR-plans in the size of the graph . A complete DR-
    plan is unique but may not be (is usually not) optimal. DR-plans
    of self-similar graphs are self-similar. More than one node (leaf)
    in a DR-plan forest may represent the same subgraph (vertex) of
    $G$.
\end{remark}

@@

\noindent
\note In Section \ref{sec:DRP} $G$ is assumed to be isostatic.

\subsection{Importance of an Optimal DR-plan}
The use of a DR-plan in designing especially self-similar qusecs is
immediately evident, as shown in Figures \ref{fig:c2c3ofk33s} and
\ref{fig:subdivision}.

In fact, knowing an optimal DR-plan is crucial for
design and analysis of even non-self-similar qusecs and their
realizations with various desirable properties.

\header{Realizing given qusecs.}
Note that this basic task
would require finding real solutions a large multivariate polynomial
system (of inequalities and equalities representing the constraints).
Without DR-planning, this requires double exponential time in the
number of variables (even if orientation type is specified). With DR-
planning, the complexity is dominated by the size of the largest
subsystem that is solved, or recombined from the solutions of its
child subsystems i.e.,the maximum fan-in occuring in a DR-plan. Thus
finding an optimal DR-plan is important for realization.

\header{Recursive block
decomposition of the rigidity matrix, stress space and external stresses.}
This can be obtained using the DR-plan we obtain a
with each block being the
rigidity matrix of an isostatic subgraph. The blocks intersect on
isostatic or trival subgraphs. This recursive block decomposition of
the rigidity matrix $R$ correspondingly recursively decomposes
the stress vectors $s$ that balance a given external stress $t$ since
$sR = t$. Moreover, the external stress vector $t$ is
recursively ``distributed'' into external stresses for each
node of the DR-plan.
When the rigidity matrix and stress
vectors contain indeterminates, this  can be used to design
underlying graph and parameters $\delta$ to obtain an optimal
distribution of stresses on the geometric primitives. When the
realization of qusecs is given, this can be used to analyze the
distribution of a given external load.
